\documentclass{article}

\usepackage{amsmath, mathrsfs, amssymb, stmaryrd, cancel, hyperref, relsize,tikz,amsthm,comment,bussproofs,xcolor,enumerate}
\usepackage{graphicx}
\usepackage{xfrac}
\hypersetup{pdfstartview={XYZ null null 1.25}}
\usepackage[all]{xy}
\usepackage[normalem]{ulem}
\usepackage{tikz-cd}


\theoremstyle{plain}
\newtheorem{theorem}{Theorem}{\bfseries}{\itshape}
\newtheorem{proposition}[theorem]{Proposition}{\bfseries}{\itshape}
\newtheorem{definition}[theorem]{Definition}{\bfseries}{\upshape}
\newtheorem{lemma}[theorem]{Lemma}{\bfseries}{\upshape}
\newtheorem{example}[theorem]{Example}{\bfseries}{\upshape}
\newtheorem{corollary}[theorem]{Corollary}{\bfseries}{\upshape}
\newtheorem{remark}[theorem]{Remark}{\bfseries}{\upshape}
\newtheorem{fact}[theorem]{Fact}{\bfseries}{\upshape}
\newtheorem{Q}[theorem]{Exercise}{\bfseries}{\upshape}

\newtheorem*{theorem*}{Theorem}

% Uncomment below to create stand-alone file
% \newcommand*{\prefix}{}

\newcommand{\sL}{\mathscr{L}}
\newcommand{\bN}{\mathbb{N}}
\newcommand{\bZ}{\mathbb{Z}}
\newcommand{\bQ}{\mathbb{Q}}
\newcommand{\bR}{\mathbb{R}}
\newcommand{\bP}{\mathbb{P}}
\newcommand{\HCF}{\mathbf{HCF}}
\newcommand{\lequiv}{\models\text{\reflectbox{$\models$}}}
\newcommand{\lra}{\leftrightarrow}
\newcommand{\ra}{\rightarrow}
\newcommand{\la}{\leftarrow}

\title{ITCS 531 \\Logic 5: Basic model theory exercises}
\author{Rob Egrot}
\date{}

\begin{document}
\maketitle
%\includecomment{comment}
\begin{Q}
Let $\phi$ be a formula where $x$ occurs free.
\begin{enumerate}[a)]
\item Write down a proof tree that shows $\forall x \phi \vdash \neg\exists x \neg\phi$. 
\item Write down a proof tree that shows $\exists x \phi \vdash \neg\forall x \neg \phi$.
\end{enumerate}
\end{Q}
\begin{comment}
\textbf{Solution:}
\begin{enumerate}[a)]
\item\begin{prooftree}
\AxiomC{$[\exists x\neg\phi]_1$}
\AxiomC{$[\neg\phi[x'/x]]_2$}
\AxiomC{$\forall x\phi$}
\RightLabel{ $\color{red}(\forall_E)$}
\UnaryInfC{$\phi[x'/x]$}
\RightLabel{ $\color{red}(\neg_E)$}
\BinaryInfC{$\bot$}
\RightLabel{ $\color{red}(\exists_E)$}
\BinaryInfC{$\bot$}
\RightLabel{ $\color{red}(\neg_I)$}
\UnaryInfC{$\neg\exists x \neg\phi$}
\end{prooftree}

\item \begin{prooftree}
\AxiomC{$\exists x \phi$}
\AxiomC{$[\forall x\neg\phi]_1$}
\LeftLabel{ $\color{red}(\forall_E)$}
\UnaryInfC{$\neg\phi[x'/x]$}
\AxiomC{$[\phi[x'/x]]_2$}
\RightLabel{ $\color{red}(\neg_E)$}
\BinaryInfC{$\bot$}
\RightLabel{ $\color{red}(\exists_E)$}
\BinaryInfC{$\bot$}
\RightLabel{ $\color{red}(\neg_I)$}
\UnaryInfC{$\neg\forall x \neg\phi$}
\end{prooftree}
\end{enumerate}
\end{comment}

\begin{Q}
Assume that lemma \ref{L:equiv} holds for first-order logic. Prove that theorems \ref{T:G1} and \ref{T:G2} are equivalent. I.e., prove the equivalence of the statements
\[\tag{$\dagger$} \text{$\Gamma$ consistent $\iff\Gamma$ is satisfiable}\]
\[\tag{$\ddagger$} \Gamma\vdash \phi\iff\Gamma\models\phi\]
\end{Q}
\begin{comment}
\textbf{Solution:}
This is similar to exercise \ref{godel_equiv}.
\begin{itemize}
\item[$\dagger\implies\ddagger$:] \begin{align*}
\Gamma\vdash \phi &\iff \Gamma\cup\{\neg\phi\}\vdash \bot\\
&\iff \Gamma\cup\{\neg\phi\}\models \bot\\
&\iff \Gamma\models \phi
\end{align*}
\item[$\ddagger\implies\dagger$:] 
\begin{align*}
\Gamma\vdash \bot &\iff \Gamma\setminus\{\phi\} \cup\{\phi\} \vdash \bot \\
&\iff \Gamma\setminus\{\phi\}\vdash \neg\phi \\
&\iff \Gamma\setminus\{\phi\}\models \neg\phi \\
&\iff\Gamma\models \bot
\end{align*}
\end{itemize}
\end{comment}


\begin{Q}
Prove that if $\Gamma$ is an $\sL$-theory then there is an $\sL$-theory $\Gamma'$ with $\Gamma\subseteq \Gamma'$ such that $\Gamma'$ is \emph{complete} (i.e. if $\phi$ is an $\sL$-sentence, then either $\phi\in\Gamma'$ or $\neg\phi\in\Gamma'$). HINT: If $\Gamma$ is consistent then it must have a .....?
\end{Q}
\begin{comment}
\textbf{Solution:}
An $\sL$-theory is a satisfiable set of $\sL$-sentences. Since $\Gamma$ is satisfiable, it must have a model. Let $A$ be a model for $\Gamma$, and let 
\[\Gamma'=\{\phi:\phi\text{ is an }\sL\text{-sentence and }A\models \phi\}.\] 
Then $\Gamma\subseteq \Gamma'$, because $A\models \Gamma$, and $\Gamma'$ is complete because every $\sL$-sentence is either true or false in $A$.
\end{comment}

\begin{Q}
Let $\Gamma$ be an $\sL$-theory, and let $\phi$ be an $\sL$-sentence. Prove that if $\Gamma\models \phi$ then $\Delta\models \phi$ for some finite $\Delta\subseteq \Gamma$.
\end{Q}
\begin{comment}
\textbf{Solution:}
Suppose $\Gamma\models \phi$. Then, by completeness we have $\Gamma\vdash \phi$. So there is a deduction tree using $\Gamma$ that proves $\phi$. As deduction trees are finite, this tree involves only a finite number of sentences from $\Gamma$. Define $\Delta$ to be the set of sentences from $\Gamma$ used in the proof of $\phi$. Then $\Delta\vdash \phi$, and so $\Delta\models \phi$ by soundness. 
\end{comment}

\begin{Q}[Compactness theorem for first-order logic] Let $\Gamma$ be a set of $\sL$-sentences. Prove that $\Gamma$ has a model if and only if every finite subset of $\Gamma$ has a model. 
\end{Q}
\begin{comment}
\textbf{Solution:}
This is very similar to exercise \ref{compactness}. First, if $\Gamma$ has a model then every subset of $\Gamma$ obviously has a model too. Conversely, suppose $\Gamma$ does \emph{not} have a model. Then $\Gamma\models \bot$. So by exercise 4 there is finite $\Delta\subseteq \Gamma$ with $\Delta\models \bot$. I.e. $\Delta$ does not have a model. 
\end{comment}
\end{document}